Can We Crunch the Social Graph on the Cheap?

### 2011-01-24 doc/hotpar2011/softxmt-hotpar2011.tex
abstract, intro

Graph analytics are very important.

Irregular applications present a challenge for parallel machines:
obtaining performance is not straightforward.

Multithreading is a technique that has been successfully used to
extract performance from these problems.

Our goal is to build a system with XMT-like properties, implemented in
software with commodity procesors

In this paper, we describe a runtime for latency tolerance. Our
analysis shows the feasibility of our approach. We end with our plans
for adding synchronization support and scaling up to multiple nodes.

  Large irregular applications such as sparse, low diameter graph problems often
  perform badly on commodity clusters. These application make many
  non-sequential accesses to global data structures with low degrees of
  locality, interacting badly with commodity processors' memory
  systems, which are optimized for local, sequential memory
  accesses. Special-purpose hardware optimized for these irregular
  applications has been built; the Cray XMT is one such machine. It
  uses multithreading to tolerate the latency of low-locality memory
  accesses, and while it exhibits excellent performance and energy
  efficiency for irregular applications, it performs badly on standard
  sequential applications \todo{Brandon: should this be more specific or worded differently? Its true sequential perform badly since switching between many threads, but also does XMT also perform  poorly (relative to alternatives) on highly parallel but regular apps with lots of temporal and spatial locality?}. We believe we can have the best of both
  worlds: we argue that modern commodity processors have many of the
  features necessary \todo{Brandon: we will have to be able to list these as features in the paper or use a different phrase; my feeling is that these are aspects of the system which are not optimized for this but are sufficient to suggest potential for enough memory concurrency} to give good performance on irregular problems,
  and that a latency-tolerant runtime designed to exploit these
  features can enable speedup for irregular applications on commodity
  processors. In this paper, we describe the key features of such a
  runtime system, and what a cluster built using our runtime might
  look like. Using a preliminary single-node implementation of our
  runtime, we obtained speedups on simple list chasing benchmarks. Our results show promise for a multi-node implementation.

\end{abstract}

@@

Graphs have proven to be a useful abstraction for solving real-world
problems in social and biological network analysis. But many graphs in
social and biological network analysis are sparse, with low diameter
and a degree distribution that follows a power law. These properties
make it difficult to take advantage of the locality assumptions built
into the memory hierarchy of modern processors, so that even simple
traversals are dominated by memory latency.

And these graphs can be large---too large for a single node. In January 2011, Facebook had 500 million active users with an average of
130 friends each \cite{Facebook:2011p91}. A natural approach would be
to try to partition the graph across the nodes in a cluster. But with
low diameter graphs, a traversal starting at any vertex is likely to
quickly leave its partition and travel over the network, incurring
high latency costs. And power law graphs have a small number of vertices
that generate significantly more work than other vertices, leading to
computational imbalance.

% Facebook's friend suggestion
% algorithm in 2010 ran on a 40-node cluster with 72GB memory per
% node. Even with that capacity, new suggestions could only be computed
% every two days \cite{Backstrom:2010p90}.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such a solution: it tries to solve the memory latency
problem through concurrency rather than caching. Each XMT processor
supports 128 hardware contexts and $\sim$1000 outstanding memory
operations and is able to switch contexts every cycle. But this
approach has a limitation: each context is significantly slower than a
modern commodity processor, yielding bad single-threaded
performance. While the XMT is fast at sparse, low-degree graph
problems, it is slow at other scientific codes that can take advantage
of commodity processors' memory hierarchies. Thus, few XMTs have been
built.

We want the best of both worlds---we want to build a system that has
good performance on both these low-locality graph codes as well as on
common scientific codes. Our approach is inspired by the XMT, but
implemented in software on commodity microprocessors. The core of our
approach is a lightweight coroutine library designed to overlap
long-latency memory prefetches with other contexts' computation.

\todo{something like: project still in early stages. this document lays out
  problems we think a system like this needs to solve, and describes
  some implementation strategies we want to explore. we have a basic
  initial implementation with some experiments that suggest there is
  promise here.}

\todo{do we want to mention hardware at this point? If so, we should
  perhaps call it a hybrid approach.}

\section{Programming model}

@@

Each coroutine executes non-preemptively \todo{seems like non-premptively isn't ``how" it executes, rather how it is scheduled, maybe say ``cooperatively"?}, utilizing the core to which
its ``parent'' thread has been assigned by the operating system until it
reaches a yield point.  Typically, the coroutine yields after
initiating a prefetch operation on data that the programmer or
compiler anticipates would probably not complete before the data is
referenced by a subsequent load operation.  In this way, the core
continues to execute instructions for other coroutines when otherwise
it would likely stall.  The programming model thus tolerates latency.

@@

\section{Approach}

Our goal is to enable high throughput on graphs having two challenging
properties. First, the graphs' combination of sparseness and low
degree means there is little locality to be exploited. Second, the
graphs are larger than a single node's memory and do not partition
well because of low diameter \todo{enough explanation now?}. We must spread the graph across multiple nodes
and send memory operations over a network. If memory requests are
uniformly distributed over the working set, a given node will make as
many accesses to remote nodes' memory as there are remote requests to the local node's memory coming in.

We would like to run as fast as current hardware will let us. Pin
bandwidth is the tightest resource: Intel's QuickPath Interconnect
(QPI) allows approximately 100 million requests per second
\todo{cite}, and recent interconnect products allow similar request
rates, with a round-trip latency of approximately 3$\mu$s
\todo{cite}. Little's law tells us then that we must support on the
order of 300 concurrent memory operations--and thus, up to this many coroutines--to
cover this latency.

These requirements drive the design of our system. There are three key
components. First, the system needs lightweight context switches (coroutines) to be able to support many contexts--thereby increasing memory concurrency available--with minimal impact on performance. Second, a memory management subsystem must support many concurrent accesses to a global address space. Third, the system must provide synchronization at a fine-granularity \todo{this last one sounds incomplete because no reason given}.

\subsection{Lightweight context switch}
\todo{can we include in the title something that means lightweight and good scheduling?}

There are three main concerns in designing the coroutine
mechanism. First, contexts must be small, so many coroutines can be
active without monopolizing the cache. Second, context switches must
be fast, so that overhead does not swamp actual work. Third, when a
coroutine yields, we must choose a suitable coroutine to run next.

@@

The choice of which coroutine to run next presents us with a
tradeoff. The simplest scheduler chooses coroutines to execute in
round-robin fashion. But this may lead to performance problems: a
coroutine might be scheduled before the event which caused it to yield (e.g., a long latency memory access)
has completed, stalling the pipeline and keeping other coroutines from
progressing. A coroutine might also be scheduled long after the event
which caused it to yield has completed; while this wouldn't block
other coroutines from proceeding, it would add unnecessary latency to
the yielding coroutine. A more intelligent scheduler might allow
coroutines to execute only when they were ready, but this could
increase the overhead of context switching. Our initial implementation
takes the simple approach, but more investigation is warranted.

@@

\subsubsection{Software management}
In order to minimize memory access overhead, the memory operation
library calls do not interact with the network interface
directly. Instead, they queue their operations to another thread, the
{\em memory manager thread}, running on another core in the CPU. This thread
translates the global address into a network destination and issues
the network operation, checks for completion, and manages the returned
data buffers.

But we have to be careful---directing memory requests from multiple
compute threads to a single memory manager thread may reduce memory
concurrency. And we have increased the overhead of a single memory
operation: what in the single node case was a simple read, is now
multiple queuing operations between threads and multiple reads and
writes over QPI to the network interface. Adding more memory manager
threads may enable more memory concurrency, but the overhead is still high \todo{why? because lots of code to run? so accel by impl in hardware. Maybe this is answered by the "what in...is now" already}

\subsubsection{Hardware management}
\todo{might want to call it accelerator throughout after saying FPGA once}
Considering this overhead, we propose hardware acceleration with a {\em memory manager
  FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
participate in the node's coherence domain. This coprocessor does no
computation---it would only manage the global memory requests of the
main CPU, performing the function of the memory manager thread
described earlier.

@@

Unfortunately, current processors do not provide sufficient hardware
resources for us to reach our concurrency target using this
approach. Each core in Intel's X5650 Xeon supports only 10 outstanding
misses from its private L2; the entire chip is limited to 32 L2 misses
to local memory and 12 L2 misses for memory reached over the QPI
bus. A simple translation of a local load to a network access would
tie up one of these slots for the entire multi-microsecond latency of
the request. To allow the amount of concurrency we need, we must bypass these structures; the CPU needs to direct the memory manager FPGA
to do work on its behalf without tying up its own resources.

To accomplish this, we encode commands in the high-order bits of
remote memory addresses. The memory manager FPGA advertises each
remote address twice \todo{this sounded like an active thing the FPGA is doing each time you make remote request, first time I read it. Consider just sticking to cmd as upper order address bits than saying advertise each address twice}. The first address is used for blocking reads and
writes---any access incurs the full latency of the remote
access. \todo{verify writes block?} The second address is used for
prefetches and asynchronous writes. These addresses differ by one
high-order bit, easily set with a mask in a macro. A read from the
prefetch address starts the network transaction, but immediately
returns a dummy value. When the value arrives, it is stored in a
temporary buffer; when the requesting coroutine is scheduled again, it
executes the blocking read, and the memory manager immediately returns
the value from the buffer \todo{is this necessarily restricted to the memory manager returning value from buffer rather than buffer in local memory?  maybe it is, since if buffer was in local mem, would have to also write a full bit or make two reads: valid bit and data}. \todo{this isn't fully baked---do we want
  to include this sort of detail?}

@@

In systems with hundreds of threads per node, lightweight
synchronization is important. While we have not implemented any
synchronization primitives yet, full-empty bits are a natural choice
given our programming model \todo{say why natural choice unless programming model section makes this point already}. Two questions then arise: at what
granularity should full-empty bits work, and what component should do
the synchronization?

@@

% 2TB global memory => 68GB of distributed FE bits.  This seems reasonable, given that a 2TB system might have 64 nodes, so ~1GB per memory manager if one manager per node.

\section{Current Implementation and Results}
\input{evaluation}

\section{Evaluation}

@@

\section{Related work}

hardware:

XMT/MTA

cyclops

niagra

why not wait for more cores?
- perhaps memory subsystems on a chip are not keeping up with the bandwidth of so many cores.

hyperthreading?

\section{Conclusion}

Large irregular problems found in areas such as social networking and
bioinformatics tend to have poor performance on commodity clusters due
to frequent low-locality memory accesses. We present a programming
model and runtime system that uses lightweight context switching to
tolerate long-latency accesses and increase performance on these
applications. Our preliminary evaluation shows that the overhead of
context switching is not prohibitive for thread counts needed for a
large-scale implementation, but memory concurrency is limited by
hardware structures in commodity processors. We describe how we plan
to address the challenges of large-scale memory concurrency and
lightweight synchronization.

>>>>

 Techniques from graph analytics have found important applications in
  areas such as social networks and bioinformatics. These irregular
  problems present a challenge for parallel machines: obtaining
  performance is not straightforward. Multithreading works well for
  such problems, but large-scale implementations of the technique like
  the Cray XMT are expensive, non-commodity machines.

  Our goal is to build a system with XMT-like properties, implemented
  with commodity processors. This paper presents a runtime for latency
  tolerance, along with our plans for scaling up to multiple
  nodes. Our analysis shows the feasibility of our approach.
\end{abstract}

@@

Techniques from graph analytics have found important applications in
areas such as social networks and bioinformatics. \todo{more!}

% In January 2011, Facebook had 500 million active users with an average
% of 130 friends each \cite{Facebook:2011p91}.
% Facebook's friend suggestion
% algorithm in 2010 ran on a 40-node cluster with 72GB memory per
% node. Even with that capacity, new suggestions could only be computed
% every two days \cite{Backstrom:2010p90}.

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. Their size keeps them from fitting in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality. Their power law distribution means
that a few vertices generate much more work than the rest, leading to
computational imbalance and even less locality. Together, these
properties lead to a difficult conclusion: each edge traversal is
likely to be a very expensive cache miss.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle. This ability comes at a cost: the
XMT is an expensive, non-commodity machine.

We want the best of both worlds. Our goal is to build a system that
has good performance on these low-locality graph codes but is
implemented with commodity processors. The core of our approach is a
lightweight coroutine library designed to overlap long-latency memory
prefetches with other contexts' computation.

In this paper, we describe our single-node runtime for latency
tolerance, along with our plans for scaling to multiple nodes.  We
evaluate the performance of our runtime using a set of list traversal
benchmarks designed to model certain worst-case behaviors of these
large graph problems.

The rest of the paper is organized as follows. We present related work
in section~\ref{sec:related}. We describe the programming model we are
implementing in section~\ref{sec:model}, and describe our approach for
implementing this model in section~\ref{sec:approach}. We evaluate our
latency-tolerant runtime in section~\ref{sec:evaluation} and conclude
in section~\ref{sec:conclusion}.

\section{Related work}
\label{sec:related}

\todo{actually write this}

Multithreading for latency tolerance has been explored in a number of directions.

Hardware approaches: The Tera MTA and Cray XMT, the MIT Alewife
project, ??? use hardware context switches to tolerate memory latency
for large systems. More recently, Simultaneous Multi-Threading,
Niagara, Cyclops, and even GPUs switch between multiple contexts to
tolerate memory latency, for memory that is hundreds of cycles of
away, rather than the thousands we must tolerate.

Mowry's SCM

Cullers' TAM

why not wait for more cores?

\section{Programming model}
\label{sec:model}

\todo{Change to discuss API:

context switches:
  yield()
  discuss coroutines

mem:
  alloc()
  read(p) 
  write(p, d)
  prefetch(p)

sync:
 {read, write} {EE, EF, FE, FF}
}

@@

Each coroutine executes non-preemptively, utilizing the core to which
its ``parent'' thread has been assigned by the operating system until it
reaches a yield point.  Typically, the coroutine yields after
initiating a prefetch operation on data that the programmer or
compiler anticipates would probably not complete before the data is
referenced by a subsequent load operation.  In this way, the core
continues to execute instructions for other coroutines when otherwise
it would likely stall.  The programming model thus tolerates latency.

@@

\section{Approach}
\label{sec:approach}

Our goal is to enable high throughput on graphs with two challenging
properties. First, the graphs' combination of sparseness and low
degree means there is little locality to be exploited. Second, the
graphs are larger than a single node's memory and do not partition
well \todo{explain?}. We must spread the graph across multiple nodes
and send memory operations over a network. If memory requests are
uniformly distributed over the working set, a given node will make as
many accesses to a distant node's memory as it does to the local node.

We would like to run as fast as current hardware will let us. Pin
bandwidth is the tightest resource: Intel's QuickPath Interconnect
(QPI) allows approximately 100 million requests per second
\todo{cite}, and recent interconnect products allow similar request
rates, with a round-trip latency of approximately 3$\mu$s
\todo{cite}. Little's law tells us then that we must support on the
order of 300 concurrent memory operations, and thus coroutines, to
cover this latency.

These requirements drive the design of our system. There are three key
components.

\subsection{Low-latency context switch}

There are three main concerns in designing the coroutine
mechanism. First, contexts must be small, so many coroutines can be
active without monopolizing the cache. Second, context switches must
be fast, so that overhead does not swamp actual work. Third, when a
coroutine yields, we must choose a good coroutine to run next.

@@

The choice of which coroutine to run next presents us with a
tradeoff. The simplest scheduler chooses coroutines to execute in
round-robin fashion. But this may lead to performance problems: a
coroutine might be scheduled before the event which caused it to yield
has completed, stalling the pipeline and keeping other coroutines from
progressing. A coroutine might also be scheduled long after the event
which caused it to yield has completed; while this wouldn't block
other coroutines from proceeding, it would add unnecessary latency to
the yielding coroutine. A more intelligent scheduler might allow
coroutines to execute only when they were ready, but this could
increase the overhead of context switching. Our initial implementation
takes the simple approach, but more investigation is warranted.

@@

In order to minimize memory access overhead, the memory operation
library calls do not interact with the network interface
directly. Instead, they queue their operations to another thread, the
{\em memory manager thread}, running on another core in the CPU. This thread
translates the global address into a network destination and issues
the network operation, checks for completion, and manages the returned
data buffers.

But we have to be careful---directing memory requests from multiple
compute threads to a single memory manager thread may reduce memory
concurrency. And we've increased the overhead of a single memory
operation---what in the single node case was a simple read, is now
multiple queuing operations between threads, and multiple reads and
writes over QPI to the network interface. Adding more memory manager
threads may enable more memory concurrency, but the overhead is still high.

Instead, we propose hardware acceleration with a {\em memory manager
  FPGA}, a coprocessor FPGA inserted into a Xeon socket so it can
participate in the node's coherence domain. This coprocessor does no
computation---it would only manage the global memory requests of the
main CPU, performing the function of the memory manager thread
described earlier.

@@

Unfortunately, current processors do not provide sufficient hardware
resources for us to reach our concurrency target using this
approach. Each core in Intel's X5650 Xeon supports only 10 outstanding
misses from its private L2; the entire chip is limited to 32 L2 misses
to local memory and 12 L2 misses for memory reached over the QPI
bus. A simple translation of a local load to a network access would
tie up one of these slots for the entire multi-microsecond latency of
the request. We must bypass these structures to allow the amount of
concurrency we need; the CPU needs to direct the memory manager FPGA
to do work on its behalf without tying up its own resources.

To accomplish this, we encode commands in the high-order bits of
remote memory addresses. The memory manager FPGA advertises each
remote address twice. The first address is used for blocking reads and
writes---any access incurs the full latency of the remote
access. \todo{verify writes block?} The second address is used for
prefetches and asynchronous writes. These addresses differ by one
high-order bit, easily set with a mask in a macro. A read from the
prefetch address starts the network transaction, but immediately
returns a dummy value. When the value arrives, it is stored in a
temporary buffer; when the requesting coroutine is scheduled again, it
executes the blocking read, and the memory manager immediately returns
the value from the buffer. \todo{this isn't fully baked---do we want
  to include this sort of detail?}

@@

In systems with hundreds of threads per node, lightweight
synchronization is important. While we have not implemented any
synchronization primitives yet, full-empty bits are a natural choice
given our programming model. Two questions then arise: at what
granularity should full-empty bits work, and what component should do
the synchronization?

@@

\section{Evaluation}
\label{sec:evaluation}

@@

\section{Conclusion}
\label{sec:conclusion}
